---
title: "Compulsory Exercise 1 in Statistical Linear Models, Spring 2021"
author: "Alexander J Ohrt"
date: "22.01.2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = "#>")
```

# Problem 1 \hspace{3mm} Bivariate normal distribution

## a)
```{r}
mu.X <- c(0,2)
Sigma.X <- matrix(c(3,1,1,3), ncol = 2)
A <- matrix(c(1/sqrt(2), 1/sqrt(2), -1/sqrt(2), 1/sqrt(2)), ncol = 2)

mu.Y <- A %*% mu.X
mu.Y

Sigma.Y <- t(A) %*% Sigma.X %*% A
Sigma.Y # Yes, the coordinates of Y are independent, since they are normally distributed and are uncorrelated. 
# This is a "special feature" of the multivariate Gaussian distribution. 
```
(Also calculated all this, in the same manner, by hand. Not sure if this should be handed in by hand or not.)
We know that the Gaussian distribution is closed under linear combinations. This means that $\mathbf{Y}$ is distributed as a bivariate Gaussian distribution, with expectation given by `mu.Y` and covariance given by `Sigma.Y`. Further, since we know that uncorrelated variables in the multivariate Gaussian distribution must be independent, this means that $Y_1$ and $Y_2$ are independent. 

## b)

(I think these facts are true, but I have no explanation yet).
Contours of the multivariate Gaussian distribution are ellipsoids. The principal axes of these ellipsoids are given by lines in the directions of the eigenvectors of the covariance matrix of the variable with the given multivariate Gaussian distribution. Furthermore, the length of the half-axes are given by the eigenvalues times their respective eigenvectors (the directions I am explaining seem correct based on the given figure, but the lengths according to the eigenvalues seem false. The relation is different in some way). 

## Problem 2 \hspace{3mm} Distributional results for $\bar{X}$ and $S^2$ for a univariate normal sample

